---
Juan Djuwadi Principal Component Analysis
title: "Principle Component Aanalysis"
output: html_document
---
## Data
The data you will be using comes from the Assistments online intelligent tutoring system (https://www.assistments.org/). It describes students working through online math problems. Each student has the following data associated with them:

- id
- prior_prob_count: How many problems a student has answered in the system prior to this session
- prior_percent_correct: The percentage of problems a student has answered correctly prior to this session
- problems_attempted: The number of problems the student has attempted in the current session
- mean_correct: The average number of correct answers a student made on their first attempt at problems in the current session
- mean_hint: The average number of hints a student asked for in the current session
- mean_attempt: The average number of attempts a student took to answer a problem in the current session
- mean_confidence: The average confidence each student has in their ability to answer the problems in the current session

## Start by uploading the data
```{r}
D1 <- read.csv("Assistments-confidence.csv")

#Removing id
D1 <- D1[,-1]

  #We won't need the id variable, so remove that.


```

## Create a correlation matrix of the relationships between the variables, including correlation coefficients for each pair of variables/features.

```{r}
#You can install the corrplot package to plot some pretty correlation matrices (sometimes called correlograms)

install.packages("corrplot")
library(corrplot)

#Generate pairwise correlations
COR <- cor(D1)

corrplot(COR, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")

#Study your correlogram image and save it, you will need it later
```

## Create a new data frame with the mean_correct variables removed

```{r}
#Removing mean_correct
D2 <- D1[,-4]

#The, scale and center your data for easier interpretation
D2 <- scale(D2, center = TRUE)
```

## Now run the PCA on the new data frame

```{r}
pca <- prcomp(D2, scale = TRUE)
```

## Although princomp does not generate the eigenvalues directly for us, we can print a list of the standard deviation of the variance accounted for by each component.

```{r}
pca$sdev

#To convert this into variance accounted for we can square it, these numbers are proportional to the eigenvalue

pca$sdev^2

#A summary of our pca will give us the proportion of variance accounted for by each component

summary(pca)

#We can look at this to get an idea of which components we should keep and which we should drop

plot(pca, type = "lines")
```

## Think about which components you would drop and make a decision

## Part II

```{r}

#Now, create a data frame of the transformed data from your pca.

D3 <- as.data.frame(pca$x)

#Attach the variable "mean_correct" from your original data frame to D3.

D4 <- cbind(D3, as.data.frame(D1$mean_correct))

#Now re-run your scatterplots and correlations between the transformed data and mean_correct. If you had dropped some components would you have lost important infomation about mean_correct?

## We may drop PC6 given the information from the plot line since PC6 shows little variation. However, an assessment of the corplot with the principal components shows that PC6 is heavily correlated with mean correct - our out come of interest. Dropping it would mean losing important information about mean correct. Dropping PC5 may be feasible since it has nearly no correlation with mean_correct as indicated by the corrplot and has relatively lower variance in comparison with the other components. 


COR2 <- cor(D4)
corrplot(COR2, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")
```
## Now print out the eigenvectors (often called loadings) for the components you generated:

```{r}
pca$rotation

#Examine the eigenvectors, notice that they are a little difficult to interpret. It is much easier to make sense of them if we make them proportional within each component

loadings <- abs(pca$rotation) #abs() will make all eigenvectors positive

sweep(loadings, 2, colSums(loadings), "/") #sweep() computes each row as a proportion of the column. (There must be a way to do this with dplyr()?)

#Calculating Each Row as a Proporton of the Column with Dplyr and Tidyr
library(dplyr)
library(tidyr)
loadings2 <- tbl_df(loadings)
Variables <- c("prior_prob_count", "prior_percent_correct", "problems_attempted", "mean_hint", "mean_attempt", "mean_confidence ") 
loadings3 <- cbind(Variables, loadings2)
loadings4 <- loadings3 %>% mutate(PC1 = PC1/sum(PC1)) %>% mutate(PC2 = PC2/sum(PC2)) %>% mutate(PC3 = PC3/sum(PC3)) %>% mutate(PC4 = PC4/sum(PC4)) %>% mutate(PC5 = PC5/sum(PC5)) %>%  mutate(PC6 = PC6/sum(PC6)) %>% 

#Now examine your components and try to come up with substantive descriptions of what some might represent?


#You can generate a biplot to help you, though these can be a bit confusing. They plot the transformed data by the first two components. Therefore, the axes represent the direction of maximum variance. Then mapped onto this point cloud are the original directions of the variables, depicted as red arrows. It is supposed to provide a visualization of which variables "go together". Variables that possibly represent the same underlying construct point in the same direction.  

biplot(pca)

#Calculate values for each student that represent these your composite variables and then create a new correlogram showing their relationship to mean_correct.

D21 <- as.matrix(D2) %*% pca$rotation
D22 <- cbind(D21, as.data.frame(D1$mean_correct))
CORCOMP <- cor(D22)
corrplot(CORCOMP, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")

```
# Part III  
## Also in this repository is a data set and codebook from Rod Martin, Patricia Puhlik-Doris, Gwen Larsen, Jeanette Gray, Kelly Weir at the University of Western Ontario about people's sense of humor. Can you perform a PCA on this data?
```{r}

#Reading the Humor Data into a Data Frame
DF <- read.csv("humor_data.csv")

#Parsing Data into Just 32 Questions, then generating a Correlation Plot to observe the relationship
DF2 <- DF[1:32]
CORDF2 <- cor(DF2)
corrplot(CORDF2, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")

#Scaling the Data 
DF3 <- scale(DF2, center = TRUE)

#Creating PCA Componenets
pca1 <- prcomp(DF3, scale = TRUE)

#Creating Sweeps and Biplot of the PCA
loadings0 <- abs(pca1$rotation)
x <- sweep(loadings0, 2, colSums(loadings0),"/")
biplot(pca1)

#We see from the biplot that the directions are generally clustered into four directions, this generally agrees with the given survey's own assessment of the questions. 


#Plotting PCA Line
plot(pca1, type = "lines")
#We see from the obervation of the plot that variables 6 through 10 can be dropped.

#We drop the all the PCA components past 6
DF4 <- as.data.frame(pca1$x)
DF5 <- DF4[1:6]


#We scale the data set and create a Corrplot
DF7 <- scale(DF5, center = TRUE)
CORDF7 <- cor(DF7)

corrplot(CORDF7, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")

#We perform PCA on the data frame after the non-explaining components are dropped.
pca2 <- prcomp(DF7, scale = TRUE)
summary(pca2)
plot(pca2, type = "lines")

#We create the Eigenvectors and create a Biplot.
loadings2 <- abs(pca2$rotation)
x <- sweep(loadings2, 2, colSums(loadings2),"/")
biplot(pca2)

```

